phoeton: Software for Objective Probabilistic Foehn Classification

### 2018-11-30 inst/draft/draft.tex
Updated draft

\textit{Note by Reto:}
The original publication (\code{FlexMix}, \citealt{gruen2008})
provides general framework for flexible mixture models for multiple
clusters. For this application we will concentrate on the special
case with only two distinct clusters ($K = 2$).

@@

The finite mixture model in its general form is given by

@@

where $\mathit{\theta}$ and $\mathit{\alpha}$ denote the parameters for the
mixture density $h()$. $\mathit{y}$ denotes the response, $\mathit{x}$ the
predictor variable (one single variable) and $\mathbf{\omega}$ the concomitant
variables (multiple variables allowed) for the probability model.

For the special case of a finite mixture model with $K=2$ Gaussian clusters
the specification as shown in Equation~\ref{eqn:flexmix-density-general} simplifies to

\begin{equation}
    h(\mathit{y} | \mathit{x}, \mathit{\omega}, \mathit{\alpha}, \mathit{\theta}) =
        \underbrace{
            (1 - \mathit{\pi}(\mathbf{\omega}, \mathit{\alpha})) \cdot
            \phi(\mathit{y} | \mathit{\theta}_1, \mathit{x})
        }_{\text{cluster 1}}
        \cdot
        \underbrace{
            \mathit{\pi}(\mathbf{\omega}, \mathit{\alpha}) \cdot \phi(\mathit{\theta}_2|\mathit{x})
        }_{\text{cluster 2}}
    \label{eqn:flexmix-density-gaussian}
\end{equation}

where $\phi()$ is the probability function of the Gaussian distribution.
The concomitant model for $\pi()$ reduces to a binomial problem as the probability
for a specific observation $y_i$ to be observed in the first cluster is the
counter probability of $y_i$ being observed in the second cluster.
Thus, for $\pi()$ any type of binomial model can be used which satisfies
that $\pi \in [0,1] \forall i=1,\dots,N$.

@@

where $\mathit{\pi}$ is the probability that an observation belongs to the
second cluster (Eqn.~\ref{eqn:flexmix-density-gaussian}).
The probabilities are modelled using multilinear regression
$\mathit{\omega}^\top \mathit{\alpha}$ with $p=1,\dots,P$ concomitant
variables $\mathit{\omega}$ times the corresponding parameters or regression
coefficients $\mathit{\alpha}$.

@@

For the special case of a two-cluster Gaussian finite mixture model the estimation
can be performed using iterative (re-)weighted empirical moments for the parameters
of the Gaussian densities ($\mathit{\theta}$) and IWLS to estimate the parameters
of the binomial logit model ($\mathit{\theta}$).
\note{This is true for $K>2$ as well, only the probability model would have to be adjusted.}

It is assumed that an unobservable latent variable $z_i \in [0,1]$ exists
for each observation $i=1, \dots N$ which defines the membership of the observation,
the probability that observation $n$ comes from the second cluster.
As the membership is unknown it has to be initially guessed:

@@

In addition, initial values for the parameters of the two Gaussian distributions
for cluster one and cluster two have to be estimated which are used as starting
values for the EM procedure. The parameters $\mathit{\theta}$ are thus initialized
as follows:

@@

\begin{algorithm}
    \caption{Parameter estimation for the two-cluster Gaussian finite mixture model
    using iterative (re-)weighted optimization. $\Phi()$ is the cumulative distribution
    function of the standard Gaussian distribution.}

@@

        \item Compute finite mixture model densities for both clusters:
            \begin{equation*}
                    d_1^{(j)} = (1 - \mathit{\pi}^{(j)}) \Phi\Big(\frac{\mathit{y} - \mu_1^{(j-1)}}{\sigma_1^{(j-1)}}\Big),
                    ~~
                    d_2^{(j)} = \mathit{\pi}^{(j)} \Phi\Big(\frac{\mathit{y} - \mu_2^{(j-1)}}{\sigma_2^{(j-1)}}\Big)
            \end{equation*}

        \item Compute posterior probabilities:
            \begin{equation*}
                \tilde{\mathit{p}}^{(j)} = \frac{d_2^{(j)}}{d_1^{(j)} + d_2^{(j)}}
            \end{equation*}

        %%%\item Compute initial log-likelihood sum:
        %%%    \begin{equation*}
        %%%        \ell^{(j)} = \sum_{i=1}^N 
        %%%        \underbrace{
        %%%            (1 - \tilde{p}_i^{(j)}) \log(d_{1,i}^{(j)})
        %%%        }_{\text{cluster 1}}
        %%%        +
        %%%        \underbrace{
        %%%            \tilde{p}_i^{(j)} \log(d_{2,i}^{(j)})
        %%%        }_{\text{cluster 2}}
        %%%        +
        %%%        \underbrace{
        %%%            (1 - \tilde{p}_i^{(j)}) \log(1 - \pi_i^{(j)}) + \tilde{p}_i^{(j)} \log(\pi_i^{(j)})
        %%%        }_{\text{concomitant model}}
        %%%        \Big)
        %%%    \end{equation*}

@@

        \item Evaluate log-likelihood given the new parameters $\mathit{\alpha}$ and $\mathit{\theta}$
            as well as the updated posterior probabilities $\mathit{\tilde{p}}$:
            \begin{equation*}
                \ell^{(j)} = \sum_{i=1}^N 
                \underbrace{
                    (1 - \tilde{p}_i^{(j)}) \log(d_{1,i}^{(j)})
                }_{\text{cluster 1}}
                +
                \underbrace{
                    \tilde{p}_i^{(j)} \log(d_{2,i}^{(j)})
                }_{\text{cluster 2}}
                +
                \underbrace{
                    (1 - \tilde{p}_i^{(j)}) \log(1 - \pi_i^{(j)}) + \tilde{p}_i^{(j)} \log(\pi_i^{(j)})
                }_{\text{concomitant model}}
                \Big)
            \end{equation*}

>>>>

\textit{Note by Reto:}
The original publication (\code{FlexMix}, \citealt{gruen2008})
provides general framework for flexible mixture models for multiple
components. For this application we will concentrate on the special
case with only two distinct components ($K = 2$).

@@

Finite mixture models are a general model class of finite
mixtures of regression models. It is assumed that the mixture
consists of a set of $K$ distinct components, each described by
a parametric distribution.
In addition, an weight is assigned to each of the $K$ classes
for each observation which describes the a-priori probability
for an observation to come from this component.
These weights can depend on multiple covariates which are
referred to as \textit{concomitant} variables.

@@

where $\mathit{\theta}$ and $\mathit{\alpha}$ denote the parameters for the
mixture density $h()$. $\mathit{y}$ denotes the response, $\mathit{x}$ the
predictor variable (one single variable) and $\mathbf{\omega}$ the concomitant
variables for the probability model.

For the special case of a finite mixture model with $K=2$ Gaussian components
the specification as shown in Equation~\ref{eqn:flexmix-density-general} simplifies to

\begin{equation}
    h(\mathit{y} | \mathit{x}, \mathit{\omega}, \mathit{\alpha}, \mathit{\theta}) =
        \underbrace{
            (1 - \mathit{\pi}(\mathbf{\omega}, \mathit{\alpha})) \cdot
            \phi(\mathit{y} | \mathit{\theta}_1, \mathit{x})
        }_{\text{component 1}}
        \cdot
        \underbrace{
            \mathit{\pi}(\mathbf{\omega}, \mathit{\alpha}) \cdot \phi(\mathit{\theta}_2|\mathit{x})
        }_{\text{component 2}}
    \label{eqn:flexmix-density-gaussian}
\end{equation}

where $\phi()$ is the probability function of the Gaussian distribution.
The concomitant model for $\pi()$ reduces to a binomial problem as the probability
for a specific observation $y_i$ to be observed in the first component is the
counter probability of $y_i$ being observed in the second component.
Thus, for $\pi()$ any type of binomial model can be used which satisfies
that $\pi \in [0,1] \forall i=1,\dots,N$.
Figure~\ref{fig:schematic_density} shows an example of a finite mixture model with
two Gaussian components with fixed $\mathit{\theta}$ coefficients but different
weights or a-priori probabilities $\tilde{p}_k$.

\begin{figure}[h]
    \centering
    \includegraphics[width = .8\textwidth]{draft_schematic_density}
    \caption{Schematic representation of the finite mixture model density
    (shaded; red dashed line) and the two Gaussian components (black/gray). 
    $\tilde{p}_1$ and $\tilde{p}_2$ are the a-prior weights for the components.
    The dashed lines always show the distribution of the component with equal
    and constant weights of $\tilde{p}_1 = \tilde{p}_2 = 0.5$.}
    \label{fig:schematic_density}
\end{figure}

@@

where $\mathit{\pi}$ is the probability that an observation belongs to the
second component (Eqn.~\ref{eqn:flexmix-density-gaussian}).
The probabilities are modelled using multilinear regression
$\mathit{\omega}^\top \mathit{\alpha}$ with $p=1,\dots,P$ concomitant
variables $\mathit{\omega}$ times the corresponding parameters or regression
coefficients $\mathit{\alpha}$.

@@

For the special case of a two-component Gaussian finite mixture model the estimation
can be performed using iterative (re-)weighted empirical moments for the parameters
of the Gaussian densities ($\mathit{\theta}$) and IWLS to estimate the parameters
of the binomial logit model ($\mathit{\theta}$).
\note{This is true for $K>2$ as well, only the probability model would have to be adjusted.}

It is assumed that an unobservable latent variable $z_i \in [0,1]$ exists
for each observation $i=1, \dots N$ which defines the membership of the observation,
the probability that observation $n$ comes from the second component.
As the membership is unknown it has to be initially guessed:

@@

In addition, initial values for the parameters of the two Gaussian distributions
for component one and component two have to be estimated which are used as starting
values for the EM procedure. The parameters $\mathit{\theta}$ are thus initialized
as follows:

@@

\begin{algorithm}
    \caption{Parameter estimation for the two-component Gaussian finite mixture model
    using iterative (re-)weighted optimization. $\Phi()$ is the cumulative distribution
    function of the Gaussian distribution.}

@@

        %%\item Compute finite mixture model densities for both components:
        %%    \begin{equation*}
        %%            d_1^{(j)} = (1 - \mathit{\pi}^{(j)}) \Phi\Big(\frac{\mathit{y} - \mu_1^{(j-1)}}{\sigma_1^{(j-1)}}\Big),
        %%            ~~
        %%            d_2^{(j)} = \mathit{\pi}^{(j)} \Phi\Big(\frac{\mathit{y} - \mu_2^{(j-1)}}{\sigma_2^{(j-1)}}\Big)
        %%    \end{equation*}

        %%\item Compute posterior probabilities:
        %%    \begin{equation*}
        %%        \tilde{\mathit{p}}^{(j)} = \frac{d_2^{(j)}}{d_1^{(j)} + d_2^{(j)}}
        %%    \end{equation*}

        \item Compute posterior probabilities:
            \begin{equation*}
                \tilde{\mathit{p}}^{(j)} = 
                \frac{
                    \mathit{\pi}^{(j-1)} \Phi(\mathit{y} | \mu_2^{(j-1)}, \sigma_2^{(j-1)})
                }{
                    (1 - \mathit{\pi}^{(j-1)}) \Phi(\mathit{y} | \mu_1^{(j-1)}, \sigma_1^{(j-1)}) +
                    \mathit{\pi}^{(j-1)} \Phi(\mathit{y} | \mu_2^{(j-1)}, \sigma_2^{(j-1)})
                }
            \end{equation*}

        %%%\item Compute initial log-likelihood sum:
        %%%    \begin{equation*}
        %%%        \ell^{(j)} = \sum_{i=1}^N 
        %%%        \underbrace{
        %%%            (1 - \tilde{p}_i^{(j)}) \log(d_{1,i}^{(j)})
        %%%        }_{\text{component 1}}
        %%%        +
        %%%        \underbrace{
        %%%            \tilde{p}_i^{(j)} \log(d_{2,i}^{(j)})
        %%%        }_{\text{component 2}}
        %%%        +
        %%%        \underbrace{
        %%%            (1 - \tilde{p}_i^{(j)}) \log(1 - \pi_i^{(j)}) + \tilde{p}_i^{(j)} \log(\pi_i^{(j)})
        %%%        }_{\text{concomitant model}}
        %%%        \Big)
        %%%    \end{equation*}
        \item Update $\mathit{\theta}$ using weighted empirical moments:
            \begin{eqnarray*}
                    \mu_1^{(j)} = \frac{\sum \mathit{x} (1 - \tilde{p}^{(j)})}{\sum (1 - \tilde{p}^{(j)})}
                    &\text{and}&
                    \sigma_1^{(j)} = \Big( \frac{\sum (x - \mu_1^{(j)})^2 (1 - \tilde{p}^{(j)})}{\sum(1 - \tilde{p}^{(j)})} \Big)^\frac{1}{2} \\
                    \mu_2^{(j)} = \frac{\sum \mathit{x} \tilde{p}^{(j)}}{\sum \tilde{p}^{(j)}}
                    &\text{and}&
                    \sigma_2^{(j)} = \Big( \frac{\sum (x - \mu_2^{(j)})^2 \tilde{p}^{(j)}}{\sum \tilde{p}^{(j)}} \Big)^\frac{1}{2} \\
            \end{eqnarray*}

@@

        \item Evaluate log-likelihood given the new parameters $\mathit{\alpha}$ and $\mathit{\theta}$
            as well as the updated posterior probabilities $\mathit{\tilde{p}}$:
            \begin{equation*}
                \ell^{(j)} = \sum_{i=1}^N \Big(
                \underbrace{
                    (1 - \tilde{p}_i^{(j)}) \log(d_{1,i}^{(j)})
                }_{\text{component 1}}
                +
                \underbrace{
                    \tilde{p}_i^{(j)} \log(d_{2,i}^{(j)})
                }_{\text{component 2}}
                +
                \underbrace{
                    (1 - \tilde{p}_i^{(j)}) \log(1 - \pi_i^{(j)}) + \tilde{p}_i^{(j)} \log(\pi_i^{(j)})
                }_{\text{concomitant model}}
                \Big)
            \end{equation*}
